---
title: "tmap - tutorial"
author: "Orlando Sabogal-Cardona"
date: "Summer 2023"
output: 
  html_notebook: 
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: true
---

# Libraries

```{r}
library(tidyverse)
library(sf)
library(tmap)
```

<br/>
<hr/>

# What it tmap

- "The short answer is that tmap is an R package for visualization spatial data. The slightly longer answer is that tmap allows users to explore, analyze, and present spatial data in a intuitive way." (Martijn Tennekes & Jakub, 2021 Nowosad)[https://r-tmap.github.io/tmap-book/intro.html]
- You can create static or interactive maps
- Conceptually, it is similar to ggplot and the *layered grammar of graphics*

# Data

```{r}
Accidents <- st_read("Montevideo_Data/Accidentes2006-2010/accidentes2006-2010.shp")
Accidents = st_set_crs(Accidents, 32721)

Stations <- st_read("Montevideo_Data/afe_estaciones/afe_estaciones.shp")
Stations <- st_set_crs(Stations, 32721)

Uruguay <- st_read("Montevideo_Data/Vectoriales_2011/ine_depto.shp")
Uruguay <- st_set_crs(Uruguay, 32721)

Bicicircuitos <- st_read(
  "Montevideo_Data/CyclingMontevideoData/v_bi_bicicircuitos/v_bi_bicicircuitos.shp")

```

# Rudiments

```{r}
tm_shape(Uruguay) + 
  tm_polygons()
```

```{r}
tm_shape(Uruguay) + 
  tm_fill()
```

```{r}
tm_shape(Uruguay) + 
  tm_fill() +
  tm_layout(bg.color = "lightblue")
```
```{r}
tm_shape(Uruguay) + 
  tm_borders(col = "black", lwd = 0.5, lty = "dashed")
```

```{r}
Montevideo <- Uruguay %>% filter(NOMBRE == "MONTEVIDEO")
```

```{r}
tm_shape(Montevideo) + 
  tm_polygons() +
  tm_shape(Accidents) + tm_bubbles()
```

```{r}
tm_shape(Montevideo) + 
  tm_polygons() +
  tm_shape(Accidents) + tm_bubbles() +
  tm_layout(frame = FALSE)
```

```{r}
MyMap <- tm_shape(Uruguay) + 
  tm_polygons(col = "deepskyblue1") 

MyMap
```

```{r}
st_bbox(Uruguay)
st_bbox(Montevideo)
```


```{r}
tm_shape(Uruguay, bbox = st_bbox(Montevideo)) + 
  tm_polygons(col = "deepskyblue1") 
```


[Nort Carolina SIDS data](https://nowosad.github.io/spData/reference/nc.sids.html)

```{r}
nc = st_read(system.file("shape/nc.shp", package="sf"))
```

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79")
```

```{r}
tm_shape(nc) + tm_fill(col = "BIR79")
```

```{r}
tm_shape(nc) + tm_fill(col = "BIR79") + tm_borders()
```

<br/>
<hr/>

# The road to a map

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79")
```

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79") + 
  tm_compass() + tm_scale_bar()
```

## Legends - tm_layout function

```{r}
tm_shape(nc) + tm_polygons(col="BIR79") + 
   tm_layout(legend.outside = TRUE, frame = FALSE)
```

```{r}
tm_shape(nc) + tm_polygons(col="BIR79") + 
    tm_layout(legend.position = c(0.2, 0.03),
              scale = 0.8,
              legend.frame = T)
```

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84") + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T)
```

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84") + 
    tm_layout(title = "THI IS MY TITLE",
              legend.position = c(0.01, 0.05),
              scale = 0.9,
              legend.frame = T) + 
   tm_add_legend(type = c("fill"),
                labels = c("Some Area", "Other Area"), 
                col = c("#66c2a5", "#e5c494"),
                title = "Other Areas") +
  tm_add_legend(type = "symbol", 
                labels = c("A point", "Other Point"), 
                col = c("#e78ac3", "grey"))
```


## Where is the south?

But coming back to our map, it needs a compass so your audience know where noth and east are.

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84") + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T) + 
    tm_compass() 
  
```


```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84") + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T) + 
    tm_compass(size = 5, type = "8star") 
  
```

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84") + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T) + 
    tm_compass(size = 5, type = "rose", lwd = 3, show.labels = 0) 
```


## The scale

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84") + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T) + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar()
```

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84") + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T) + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0, 50, 100, 150, 200), text.size = 2, position = c(0.06, 0))
```

Documentation for the [tm_layout() function](https://www.rdocumentation.org/packages/tmap/versions/2.2/topics/tm_layout)
Documentation for the [tm_compass() function](https://www.rdocumentation.org/packages/tmap/versions/2.2/topics/tm_compass)
Documentation for the [tm_scale_bar() function](https://www.rdocumentation.org/packages/tmap/versions/2.2/topics/tm_scale_bar)

```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84") + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T) + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0, 50, 100, 150, 200), text.size = 1, position = c(0.06, 0),
                 lwd = 5, color.dark = "darkgreen", color.light = "blue")
```

```{r}
MyMap <- tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84") + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T,
              bg.color = "lightblue") + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0, 50, 100, 150, 200), text.size = 1, position = c(0.06, 0),
                 lwd = 3)

```

```{r}
MyMap
```

## Mode: View

```{r}
tmap_mode("view")
MyMap
```


```{r}
tmap_mode("plot")
```

## Colors

```{r}
tm_shape(nc) + tm_polygons(col="BIR79", title = "Births 1979 - 84", palette = "Blues") + 
    tm_layout(legend.position = c(0.01,0.65),
              scale=0.65,
              legend.frame = T,
              bg.color = "lightblue") + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0,50,100,150,200), text.size = 1, position = c(0.06,0),
                 lwd = 3)

```

```{r}
tm_shape(nc) + 
    tm_polygons(col="BIR79", title = "Births 1979 - 84", palette = "Blues",  n = 4) + 
    tm_layout(legend.position = c(0.01,0.65),
              scale=0.65,
              legend.frame = T,
              bg.color = "lightblue") + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0,50,100,150,200), text.size = 1, position = c(0.06,0),
                 lwd = 3)
```

```{r}
tm_shape(nc) + 
    tm_polygons(col="BIR79", title = "Births 1979 - 84", palette = "Blues", 
                style = "pretty") + 
    tm_layout(legend.position = c(0.01,0.65),
              scale=0.65,
              legend.frame = T,
              bg.color = "lightblue") + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0,50,100,150,200), text.size = 1, position = c(0.06,0),
                 lwd = 3)
```

```{r}
library(RColorBrewer)
display.brewer.all()
```


```{r}
tm_shape(nc) + tm_polygons(col="BIR79", title = "Births 1979 - 84", palette = "YlGnBu") + 
    tm_layout(legend.position = c(0.01,0.65),
              scale=0.65,
              legend.frame = T,
              bg.color = "lightblue") + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0,50,100,150,200),size = 1, position = c(0.06,0),
                 lwd = 3)
```

```{r}
tm_shape(nc) + tm_polygons(col="BIR79", title = "Births 1979 - 84", palette = "-YlGnBu") + 
    tm_layout(legend.position = c(0.01,0.65),
              scale=0.65,
              legend.frame = T,
              bg.color = "lightblue") + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0,50,100,150,200), text.size = 1, position = c(0.06,0),
                 lwd = 3)
```


```{r}
summary(nc$BIR79)
```


```{r}
nc %>% ggplot(aes(x=BIR79)) + geom_histogram()
```



```{r}
Breaks <- c(0, 2000, 4000, 6000, 31000)
Labels <- c("0 - 2000", "2000 - 4000", "4000 - 6000", ">6000")
```


```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84", palette = "YlGnBu",
                           breaks = Breaks, labels = Labels) + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T,
              bg.color = "lightblue") + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0, 50, 100, 150, 200), text.size = 1, position = c(0.06, 0),
                 lwd = 3)
```

```{r}
MyPalette <- c("#f2f0f7", "#cbc9e2", "#9e9ac8", "#6a51a3")
```


```{r}
tm_shape(nc) + tm_polygons(col = "BIR79", title = "Births 1979 - 84", palette = MyPalette,
                           breaks = Breaks, labels = Labels) + 
    tm_layout(legend.position = c(0.01, 0.65),
              scale = 0.65,
              legend.frame = T,
              bg.color = "lightblue") + 
    tm_compass(size = 5, type = "rose", lwd = 3) + 
    tm_scale_bar(breaks = c(0, 50, 100, 150, 200), text.size = 1, position = c(0.06, 0),
                 lwd = 3)
```

<br/>
<hr/>

# Auxiliar map

## Alternative A

```{r}
MontevideoMap <- tm_shape(Montevideo) + 
  tm_polygons() +
  tm_shape(Accidents) + tm_bubbles()

UruguayMap <- tm_shape(Uruguay) + 
  tm_polygons()
```

```{r}
library(grid)
```

```{r}
MontevideoMap
print(UruguayMap, vp = viewport(0.17, 0.8, width = 0.25, height = 0.25))
```

```{r}
MontevideoMap <- tm_shape(Montevideo) + 
  tm_polygons(col = "deepskyblue1") 


MontevideoNeigh <- Uruguay %>% filter(NOMBRE %in% c("SAN JOSE", "CANELONES", "MONTEVIDEO"))

AuxMap <- tm_shape(MontevideoNeigh) + tm_polygons(col = "green")
```

```{r}
MontevideoMap
print(AuxMap, vp = viewport(0.17, 0.8, width = 0.2, height = 0.2))
```


Our map needs some context!

```{r}
MontevideoBox <- st_bbox(Montevideo) %>% st_as_sfc() #Why not st_as_sf?
```

```{r}
tm_shape(MontevideoBox) + tm_polygons(col = "white", border.col = "blue", lwd = 4)
#plot(st_geometry(MontevideoBox))
```


```{r}

AuxMap <- tm_shape(MontevideoNeigh) + tm_polygons(col = "green") + 
  tm_shape(MontevideoBox) + tm_polygons(border.col = "red", lwd = 2.5, alpha = 0.3)
  

MontevideoMap
print(AuxMap, vp = viewport(0.17, 0.8, width = 0.2, height = 0.2))
```

## Alternative B

[See this link](https://github.com/r-tmap/tmap/issues/541)


<br/>
<hr/>

# Faceted maps

```{r}
tm_shape(Montevideo) + tm_polygons() +
  tm_shape(Accidents) + tm_dots(alpha = 0.5) + tm_facets(by = "ANIO")
```

```{r}
tm_shape(Montevideo) + tm_polygons() +
  tm_shape(Accidents) + tm_dots(alpha = 0.5) + tm_facets(by = "ANIO", nrow = 1)
```

```{r}
tm_shape(Montevideo) + tm_polygons() +
  tm_shape(Accidents) + tm_dots(alpha = 0.5) + tm_facets(by = "ANIO", ncol = 1)
```

```{r}
Accidents2006 <- Accidents %>% filter(ANIO == "2006")
Accidents2010 <- Accidents %>% filter(ANIO == "2010")

Complements <- tm_compass() + tm_scale_bar()

Map2006 <- tm_shape(Accidents2006) + tm_dots() + Complements
Map2010 <- tm_shape(Accidents2010) + tm_dots() + Complements
```

```{r}
tmap_arrange(Map2006, Map2010)
```


# Export

# Animated maps

See [section 8.5 of the Gecomputation with R book](https://bookdown.org/robinlovelace/geocompr/adv-map.html#static-maps)

# Recommended readings

- tmap book: [Elegant and informative maps with tmap](https://r-tmap.github.io/tmap-book/)
- Chapter 8 of [Geocomputation with R book](https://bookdown.org/robinlovelace/geocompr/)
- [tmap: get started - vignettes](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)

